### 2017-08-25 reports/thesis/book.tex
Started on EM in thesis.

\begin{document}

@@

The compound Poisson can be formally written. Let $Y$ be the number of photons detected in a pixel for some time exposure $\tau$,
\begin{equation*}
Y\sim\poisson(\lambda)
\end{equation*}
and
\begin{equation*}
\prob(Y=y)=\euler^{-\lambda}\frac{\lambda^y}{y!} \quad \text{for }y=1,2,3,\dotdotdot
\end{equation*}
where $\lambda>0$ is the Poisson rate parameter. Let $U_i$ be the energy of the $i$th photon in a pixel for i.i.d.~$i=1,2,3,\dotdotdot$ with p.d.f.~$p_U(u)$. Let $X$ be the greyvalue of a pixel and is an energy integrating detector, then
\begin{equation}
X|Y = \sum_{i=1}^{Y}U_i+\kappa
\label{eq:compoundPoisson_X|Y}
\end{equation}
where $\kappa$ is some offset. For now let $\kappa=0$, then the marginal $X$ has the compound Poisson distribution. The p.d.f.~of $X$ can be obtained by marginalising the joint p.d.f.
\begin{equation*}
p_X(x)=\sum_{y=0}^\infty p_{X|Y}(x|y)\prob(Y=y) \quad\text{for }x\geqslant 0\ .
\end{equation*}
However it should be noted that $X=0$ if and only if $Y=0$ with probability
\begin{equation*}
\prob(X=0)=\euler^{-\lambda} \ .
\end{equation*}
Formally $X$ has probability mass at $X=0$ and probability density at $X>0$, then 
\begin{equation}
p_X(x) = \delta(x) \euler^{-\lambda} + \sum_{y=1}^\infty p_{X|Y}(x|y)\euler^{-\lambda}\frac{\lambda^y}{y!} \quad\text{for }x\geqslant 0
\end{equation}
where $\delta(x)$ is the Dirac delta function. It is quite often the case this cannot be written in closed form.

@@

The p.d.f.~can be shown to be
\begin{equation}
p_X(x) = 
\delta(x) \euler^{-\lambda}
+
\sum_{y=1}^{\infty}\frac{\beta^{y\alpha}}{\Gamma(y\alpha)}x^{y\alpha-1}\euler^{-\beta x}\euler^{-\lambda}\frac{\lambda^y}{y!}
\quad\text{for }x\geqslant 0
\label{eq:compoundPoisson_pdf}
\end{equation}
and it is well known the infinite sum cannot be written in closed form. There are a number of approximations or computational methods to evaluate the p.d.f.~such as Fourier inverting the characteristic function \citep{dunn2008evaluation}, using the saddlepoint approximation \citep{daniels1954saddlepoint} or cleverly sum over certain terms in the infinite sum \citep{dunn2005series}. Of course direction simulation of a compound Poisson-Gamma random variable is incredibly easy if $U$ can be simulated. The p.d.f.~then can be estimated using these simulated samples.

@@

\section{Density Evaluation}
Suppose $X\sim\CPoisson(\lambda,\alpha,\beta)$ and $\lambda,\alpha,\beta>0$ then the p.d.f.~of $X$ is
\begin{equation}
p_X(x) = 
\delta(x) \euler^{-\lambda}
+
\euler^{-\beta x-\lambda}\frac{1}{x}\sum_{y=1}^{\infty}W_y
\quad\text{for }x\geqslant 0
\end{equation}
where
\begin{equation}
W_y = \frac{\beta^{y\alpha}\lambda^yx^{y\alpha}}{\Gamma(y\alpha)y!} \ .
\end{equation}
The evaluation of this p.d.f.~is difficult because the sum of $W_y$ cannot be simplified. However the m.g.f.~of $X$ can be written down in closed form
\begin{equation}
M(\theta)=\exp\left[\lambda\left(\left(\frac{\beta}{\beta-\theta}\right)^{\alpha}-1\right)\right] \ .
\end{equation}
The m.g.f.~allows the study of limiting distributions of the compound Poisson-Gamma distribution. These limiting distributions then can be used to approximate the p.d.f.~of a compound Poisson-Gamma random variable.

@@

For large $\lambda$
\begin{equation}
\lim_{\lambda\rightarrow\infty}M_Z(\theta)=\exp\left[\frac{\theta^2}{2}\right]
\end{equation}
which is the same as the m.g.f.~of a standard Normal variable, therefore
\begin{equation}
\lim_{\lambda\rightarrow\infty}Z\sim\normal(0,1) \ .
\end{equation}
This should make sense as for high $\lambda$, the Poisson random variable has a high expectation, increasing the number of Gamma random variables in a summation. Increasing the number of terms in a summation will triggle the central limit theorem.

@@

>>>>

\begin{document}\sloppy

@@

The compound Poisson can be formally written. Let $Y$ be the number of photons detected in a pixel for some time exposure $\tau$,
\begin{equation*}
Y\sim\poisson(\lambda)
\end{equation*}
and
\begin{equation*}
\prob(Y=y)=\euler^{-\lambda}\frac{\lambda^y}{y!} \quad \text{for }y=1,2,3,\dotdotdot
\end{equation*}
where $\lambda>0$ is the Poisson rate parameter. Let $U_i$ be the energy of the $i$th photon in a pixel for i.i.d.~$i=1,2,3,\dotdotdot$ with p.d.f.~$p_U(u)$. Let $X$ be the greyvalue of a pixel and is an energy integrating detector, then
\begin{equation}
X|Y = \sum_{i=1}^{Y}U_i+\kappa
\label{eq:compoundPoisson_X|Y}
\end{equation}
where $\kappa$ is some offset. For now let $\kappa=0$, then the marginal $X$ has the compound Poisson distribution. The p.d.f.~of $X$ can be obtained by marginalising the joint p.d.f.
\begin{equation*}
p_X(x)=\sum_{y=0}^\infty p_{X|Y}(x|y)\prob(Y=y) \quad\text{for }x\geqslant 0\ .
\end{equation*}
However it should be noted that $X=0$ if and only if $Y=0$ with probability
\begin{equation*}
\prob(X=0)=\euler^{-\lambda} \ .
\end{equation*}
Formally $X$ has probability mass at $X=0$ and probability density at $X>0$, then 
\begin{equation}
p_X(x) = 
\begin{cases}
\delta(x) \euler^{-\lambda}  & \text{ for } x=0 \\ 
\sum_{y=1}^\infty p_{X|Y}(x|y)\euler^{-\lambda}\frac{\lambda^y}{y!} \quad\text{for } & \text{ for } x>0
\end{cases}
\end{equation}
where $\delta(x)$ is the Dirac delta function. It is quite often the case this cannot be written in closed form.

@@

The p.d.f.~can be shown to be
\begin{equation}
p_X(x) = 
\begin{cases}
\delta(x) \euler^{-\lambda} & \text{ for } x=x \\ 
\sum_{y=1}^{\infty}\frac{\beta^{y\alpha}}{\Gamma(y\alpha)}x^{y\alpha-1}\euler^{-\beta x}\euler^{-\lambda}\frac{\lambda^y}{y!} & \text{ for } x>0
\end{cases}
\label{eq:compoundPoisson_pdf}
\end{equation}
and it is well known the infinite sum cannot be written in closed form. There are a number of approximations or computational methods to evaluate the p.d.f.~such as Fourier inverting the characteristic function \citep{dunn2008evaluation}, using the saddlepoint approximation \citep{daniels1954saddlepoint} or cleverly sum over certain terms in the infinite sum \citep{dunn2005series}. Of course direction simulation of a compound Poisson-Gamma random variable is incredibly easy if $U$ can be simulated. The p.d.f.~then can be estimated using these simulated samples.

@@

\section{Density Evaluation}
Suppose $X\sim\CPoisson(\lambda,\alpha,\beta)$ and $\lambda,\alpha,\beta>0$ then the p.d.f.~of $X$ is
\begin{equation}
p_X(x) = 
\begin{cases}
\delta(x) \euler^{-\lambda} & \text{ for } x=0 \\ 
\euler^{-\beta x-\lambda}\frac{1}{x}\sum_{y=1}^{\infty}W_y & \text{ for } x>0 
\end{cases}
\end{equation}
where
\begin{equation}
W_y = \frac{\beta^{y\alpha}\lambda^yx^{y\alpha}}{\Gamma(y\alpha)y!} \ .
\label{eq:compoundPoisson_w}
\end{equation}
The evaluation of this p.d.f.~is difficult because the sum of $W_y$ cannot be simplified. However the m.g.f.~of $X$ can be written down in closed form
\begin{equation}
M(\theta)=\exp\left[\lambda\left(\left(\frac{\beta}{\beta-\theta}\right)^{\alpha}-1\right)\right] \ .
\end{equation}
The m.g.f.~allows the study of limiting distributions of the compound Poisson-Gamma distribution. These limiting distributions then can be used to approximate the p.d.f.~of a compound Poisson-Gamma random variable.

@@

For large $\lambda$
\begin{equation}
\lim_{\lambda\rightarrow\infty}M_Z(\theta)=\exp\left[\frac{\theta^2}{2}\right]
\end{equation}
which is the same as the m.g.f.~of a standard Normal variable, therefore
\begin{equation}
\lim_{\lambda\rightarrow\infty}Z\sim\normal(0,1) \ .
\end{equation}
This should make sense as for high $\lambda$, the Poisson random variable has a high expectation, increasing the number of Gamma random variables in a summation. Increasing the number of terms in a summation will trigger the central limit theorem.

@@

\section{EM Algorithm}

The EM algorithm \citep{dempster1977maximum} will be proposed here as a method to estimate the parameters of a $X\sim\CPoisson(\lambda,\alpha,\beta)$ random variable given a sample of measurements of it $\left\{x_1,x_2,x_3,\dotdotdot,x_n\right\}$.

A popular method for parameter estimation is the maximum log likelihood method. Let $\widehat{\lambda}$, $\widehat{\alpha}$ and $\widehat{\beta}$ be estimators of $\lambda$, $\alpha$ and $\beta$ respectively. The log likelihood is defined to be
\begin{equation*}
\ln L(\lambda,\alpha,\beta;X) = \sum_{i=1}^n\left[
\mathbb{I}(x_i=0)\ln \prob(X=x_i)
+\mathbb{I}(x_i>0)\ln p_X(x_i)
\right]
\end{equation*}
\begin{multline}
\ln L(\lambda,\alpha,\beta;X) = \sum_{i=1}^n\left[
-\mathbb{I}(x_i=0)\lambda
\vphantom{\sum_i^i}\right.\\\left.
+\mathbb{I}(x_i>0)\left(
	-\beta x_i-\lambda-\ln x -\ln\sum_{y=1}^\infty W_y
\right)
\right]
\end{multline}
where $W_y=W_y(\lambda,\alpha,\beta)$ was defined in Equation \eqref{eq:compoundPoisson_w}. $\widehat{\lambda}$, $\widehat{\alpha}$ and $\widehat{\beta}$ are values of $\lambda$, $\alpha$ and $\beta$ which jointly maximises the log likelihood. The gradient of the log likelihood can be evaluated \citep{dunn2005series} and optimisation can be done using gradient methods.

The EM algorithm \citep{dempster1977maximum} instead treats $Y\sim\poisson(\lambda)$ as a latent variable. Let $\left\{y_1,y_2,y_3,\dotdotdot, y_n\right\}$ be measurements of $Y$ and $\left\{x_1,x_2,x_3,\dotdotdot, x_n\right\}$ be measurements of $X|Y$. Define the joint log likelihood to be
\begin{multline*}
\ln L(\lambda,\alpha,\beta;X,Y)
=
\sum_{i=1}^n
\left[
	\mathbb{I}(y_i=0)
	\ln
	\prob(Y=0)
	\right.\\\left.
	+
	\mathbb{I}(y_i>0)
	\ln
	\left[
	p_{X|Y}(x_i|y_i)\prob(Y=y_i)
	\right]
\right]
\end{multline*}
so that
\begin{multline}
\ln L(\lambda,\alpha,\beta;X,Y)
=
\sum_{i=1}^n
\left[
	-\mathbb{I}(y_i=0)
	\lambda
	\right.\\\left.+
	\mathbb{I}(y_i>0)
	\left(
	y_i\alpha\ln\beta-\ln\Gamma(y_i\alpha)+(y_i\alpha-1)\ln x_i - \beta x_i
	\right.\right.\\\left.\left.	
	- \lambda + y_i \ln \lambda - \ln(y!)
	\right)
\right] \ .
\end{multline}
The estimators are found by optimising the joint log likelihood. This is done iteratively by estimating the $Y$s given the $X$s and parameters (E step), followed by estimating the parameters given the $Y$s and $X$s (M step) until some convergence conditions are met.

It will be shown that using some approximations, the E step and M step can be implemented. The Cram\'er-Rao lower bound \citep{rao1945information} \citep{cramer1946mathematical} can also be found by using a few approximations. However simulations show that these estimators struggle for high $\lambda$ and it will be shown that this will be the case for all estimators.

\subsection{E Step}

\subsection{M Step}

\subsection{Estimation Error}

\subsection{Simulations}

\subsection{Failure Cases}
